\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-07-29 paper/colm_camera/colm2024_conference.tex
Camera-ready draft

\newcommand{\SM}[1]{\textcolor{blue}{[SM: #1]}}
\newcommand{\changes}[1]{\textcolor{blue}{#1}}

@@

\author{Antiquus S.~Hippocampus, Natalia Cerebro \& Amelie P. Amygdale \thanks{ Use footnote for providing further information
about author (webpage, alternative address)---\emph{not} for acknowledging
funding agencies.  Funding acknowledgements go at the end of the paper.} \\
Department of Computer Science\\
Cranberry-Lemon University\\
Pittsburgh, PA 15213, USA \\
\texttt{\{hippo,brain,jen\}@cs.cranberry-lemon.edu} \\
\And
Ji Q. Ren \& Yevgeny LeNet \\
Department of Computational Neuroscience \\
University of the Witwatersrand \\
Joburg, South Africa \\
\texttt{\{robot,net\}@wits.ac.za} \\
\AND
Coauthor \\
Affiliation \\
Address \\
\texttt{email}
}

@@

% \colmfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}

@@

In this paper, we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training),
% \footnote{Technically, we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), 
a method that connects, for the first time, astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals written by astronomers and images corresponding to downstream observations imaged by telescopes. 
%
Concretely, we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract pairs results in meaningful joint representations through quantitative and qualitative evaluation tests.
%
Our method opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm. A high-level overview of the method is shown in Fig.~\ref{fig:overview}.
%

@@

   In total, 31,859 images corresponding to 4,438 abstracts are included in the fine-tuning dataset.
   %
   3,194 images are held out for validation, with no abstract being common between training and validation sets in order to ensure an independent set of image-text pairs for evaluation. The held out images correspond to 429 unique abstracts.  % Double check numbers in final version.

@@

   While the base model shows some signs of meaningful retrieval (e.g., the image of Jupiter in the first row of Tab.~\ref{tab:tti_base}), it is challenging to discern meaningful, strong associations between the retrieved images and corresponding query.

@@

   Note that we chose to illustrate qualitative performance on image retrieval using the model fine-tuned on summarized abstracts, rather than raw abstracts. 
  %  We show analogous results for the model fine-tuned on raw abstracts in App.~\ref{app:eval_raw}. 
   Although the two models show very similar quantitative performance on retrieval metrics (as shown in Fig.~\ref{fig:sim_valtrain}), they exhibit characteristically different behaviors in terms of images retrieved, with the summary fine-tuned models generally retrieving images that look more visually ``relevant'' to a domain expert. We emphasize that for scientific usefulness, the goal is not necessarily to correctly retrieve the most relevant objects, but rather to identify a diverse set of interesting candidates for manual follow-up and further analysis.  % ; both models are seen to perform sensibly, even if differently, in this regard.

@@

@@

\section{Ethics Statement}
This work relies on using abstracts from successful \hubble Space Telescope observing proposals as part of a dataset for training and evaluating machine learning models. While these abstracts are publicly available, the authors likely did not anticipate their text being used in this manner, raising questions around consent, attribution, and appropriate use of data. Since this research intends to develop methods to aid astronomical research and does not use sensitive personal information or target commercial gain, we believe that the scientific benefits outweigh the potential concerns in this case, while acknowledging good-faith arguments to the contrary. As the use of foundation models in the sciences increases, it will be important for the community to consider norms and guidelines around the appropriate use and attribution of various data sources for model training and evaluation, including qualitative textual data, to ensure transparency and maintain trust.

@@

>>>>

\newcommand{\SM}[1]{\textcolor{blue}{[SM: #1]}}
\newcommand{\changes}[1]{\textcolor{red}{#1}}

@@

\author{Siddharth Mishra-Sharma, Yiding Song, \& Jesse Thaler \\
MIT \& IAIFI \\
\texttt{\{smsharma,ydsong,jthaler\}@mit.edu}
}

%\author{Siddharth Mishra-Sharma \\
%The NSF AI Institute for Artificial Intelligence and Fundamental Interactions \\
%Center for Theoretical Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
%Department of Physics, Harvard University, Cambridge, MA 02138, USA \\
%\texttt{\href{mailto:smsharma@mit.edu}{smsharma@mit.edu}}
%\And
%Yiding Song \\
%The NSF AI Institute for Artificial Intelligence and Fundamental Interactions \\
%Department of Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
%\texttt{\href{mailto:ydsong@mit.edu}{ydsong@mit.edu}}
%\And
%Jesse Thaler \\
%The NSF AI Institute for Artificial Intelligence and Fundamental Interactions \\
%Center for Theoretical Physics, Massachusetts Institute of Technology, Cambridge, MA 02139, USA \\
%\texttt{\href{mailto:jthaler@mit.edu}{jthaler@mit.edu}}
%}

@@

\colmfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}

@@

In this paper, we describe \text{PAPERCLIP} (Proposal Abstracts Provide an Effective Representation for Contrastive Language-Image Pre-training),
% \footnote{Technically, we fine tune rather than pre train, but ``PAPERCLIFT'' was rejected by the senior author of this paper.}), 
a method that connects, for the first time, astronomical image observations with natural language by leveraging the association between abstracts of successful observing proposals written by astronomers and images corresponding to downstream observations imaged by telescopes. 
%
\changes{This approach demonstrates the potential of adapting generalist multi-modal foundation models to astronomy, complementing task-specific models in the domain by providing a flexible, language-based interface for interacting with observational data.}
%
Concretely, we showcase the method using observations imaged by the \hubble Space Telescope (HST).
%
We show that fine-tuning a pre-trained CLIP ~\citep[Contrastive Language-Image Pre-training; ][]{radford2021learning} image-text model on observation-abstract pairs results in meaningful joint representations through quantitative and qualitative evaluation tests.
%
Our method opens up the possibility of interacting with astronomical survey data using free-form natural language as an interface, which is a cornerstone of the success of the modern foundation model paradigm. A high-level overview of the method is shown in Fig.~\ref{fig:overview}.
%

@@

   In total, 31,859 images corresponding to 4,438 abstracts are included in the fine-tuning dataset.
   %
   3,194 images are held out for validation, with no abstract being common between training and validation sets in order to ensure an independent set of image-text pairs for evaluation. The held out images correspond to 429 unique abstracts.  % Double check numbers in final version.
   %
   \changes{Due to practical limitations associated with the small size of the fine-tuning dataset, we did not use different datasets for validation and testing, deeming the current approach sufficient for a proof-of-principle exposition.}

@@

   While the base model shows some signs of meaningful retrieval (e.g., the image of Jupiter in the first row of Tab.~\ref{tab:tti_base}), 
  %  it is challenging to discern meaningful, strong associations between the retrieved images and corresponding query.
   \changes{associations between the retrieved images and corresponding queries are not consistent.}

@@

   \changes{We also evaluate the observation retrieval task more quantitatively. We design a prompt which lets us evaluate whether the abstract corresponding to a retrieved observation is relevant or not, with the output constrained to be a boolean using \package{Outlines}. We then evaluated this prompt, for the base as well as fine-tuned models, on the top 10 closest images by cosine similarity returned for 10 different queries. 38\% and 77\% of the retrieved observations are deemed relevant when using the base and fine-tuned models, respectively. The fine-tuned model is thus significantly more likely to return images relevant to the query. The prompt and curated queries for this test are described in App.~\ref{app:quanteval_prompt}.}

   Note that we chose to illustrate qualitative performance on image retrieval using the model fine-tuned on summarized abstracts, rather than raw abstracts. 
  %  We show analogous results for the model fine-tuned on raw abstracts in App.~\ref{app:eval_raw}. 
   Although the two models show very similar quantitative performance on retrieval metrics (as shown in Fig.~\ref{fig:sim_valtrain}), they exhibit characteristically different behaviors in terms of images retrieved, with the summary fine-tuned models generally retrieving images that look more visually ``relevant'' to a domain expert. We emphasize that for scientific usefulness, the goal is not necessarily to correctly retrieve the most relevant objects, but rather to identify a diverse set of interesting candidates for manual follow-up and further analysis.  % ; both models are seen to perform sensibly, even if differently, in this regard.
   %
   \changes{By diverse, we mean that retrieved observations may contain different types of objects or phenomena, which may be relevant to the query in distinct ways.}

@@

We thank Michael Brenner, Fran\c{c}ois Lanusse, and Julian Mu\~{n}oz for helpful conversations.
%
This work is supported by the National Science Foundation under Cooperative Agreement PHY-2019786 (The NSF AI Institute for Artificial Intelligence and Fundamental Interactions, \url{http://iaifi.org/}).
%
This material is based upon work supported by the U.S. Department of Energy, Office of Science, Office of High Energy Physics of U.S. Department of Energy under grant Contract Number  DE-SC0012567. 
%
YS was supported by the Research Science Institute (RSI) program at MIT.
%
This research was supported by an award from Google,  ``Interpretation of Multimodal Images from Astronomy''.
%
This research was supported by the Munich Institute for Astro-, Particle and BioPhysics (MIAPbP), which is funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) under Germany's Excellence Strategy – EXC-2094 – 390783311.
%
The computations in this paper were run on the FASRC Cannon cluster supported by the FAS Division of Science Research Computing Group at Harvard University.

@@

 \subsubsection*{Reproducibility Statement}

 Code used to reproduce the results in this work is available at \url{https://github.com/smsharma/PAPERCLIP-Hubble/tree/main}.

\subsubsection*{Ethics Statement}
This work relies on using abstracts from successful \hubble Space Telescope observing proposals as part of a dataset for training and evaluating machine learning models. While these abstracts are publicly available, the authors likely did not anticipate their text being used in this manner, raising questions around consent, attribution, and appropriate use of data. Since this research intends to develop methods to aid astronomical research and does not use sensitive personal information or target commercial gain, we believe that the scientific benefits outweigh the potential concerns in this case, while acknowledging good-faith arguments to the contrary. As the use of foundation models in the sciences increases, it will be important for the community to consider norms and guidelines around the appropriate use and attribution of various data sources for model training and evaluation, including qualitative textual data, to ensure transparency and maintain trust.

@@

\subsection{Prompt Used for Quantitative Evaluation of Observation Retrieval}
\label{app:quanteval_prompt}

\changes{The following prompt was used to evaluate the relevance of abstracts corresponding to retrieved images to a query, when quantitatively assessing the observation retrieval task.}

@outlines.prompt
def prompt_fn(abstract, query):
     """[INST]
You are an expert astrophysicist, with broad expertise across observational and theoretical astrophysics.

Abstract: "{{abstract}}"
Query: "{{query}}"

The above is an abstract for a proposed observation taken by the Hubble Space Telescope (labeled "Abstract"), and an object or concept (labeled "Query").

Could the observations corresponding to the abstract contain the query? Be precise, and do not contain related concepts or objects.

Your response should be either True or False. Only return True if the query is closely related to the abstract, and the downstream observation could be relevant to the query.
[/INST]
"""
\end{lstlisting}

\changes{The queries used in the evaluation were \texttt{["globular cluster", "dwarf galaxy", "SN1987A", "strong lensing", "galaxy clusters", "interstellar medium", "dark matter", "spiral galaxies", "lyman alpha", "comets"].}}
